E-Graphs With Bindings

### 2025-01-22 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
fixed some references and figs

% for quotienting
\usepackage{xfrac}

@@

Another distinct challenge is presented by $\beta$-reduction, which is handled using \emph{explicit substitutions}. 
However, calculi of explicit substitutions have many technical subtleties which also directly impact how equivalence is defined, for instance \emph{graph equivalence} instituted between terms which have non-interfering explicit substitutions applied in different orders~\cite{doi:10.1145/2535838.2535886}. 
These difficulties are not handled in \emph{loc. cit.}.

@@

The general background literature for string diagrams is summarised in~\cite{https://arxiv.org/abs/2305.08768}.
The particular context of string diagrams with bindings is given in~\cite{https://arxiv.org/abs/2305.18945}, while string diagrams for e-graphs without bindings are defined in~\cite{https://arxiv.org/abs/2406.15882}.

String diagrams provide a geometric or topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out. 
Key structures in monoidal categories---such as the monoidal product, tensor unit, braiding, and dualities---are represented through operations like placing, weaving, or bending strings. 
This visual approach makes complex operations such as the trace intuitive and reveals insights sometimes obscured by symbolic notation. 
Originating from notations by Hotz~\cite{??} (for automata) and Penrose~\cite{??} (for tensor networks), string diagrams generalise to any monoidal category and even bicategories, offering a dual perspective to commutative diagrams through Poincaré duality.

As described above, string diagrams are commonly an informal yet rigorous two-dimensional syntax for the internal language of a category. 
They can be formalised by representing them as graphs or hypergraphs~\cite{https://dl.acm.org/doi/full/10.1145/3502719}, which has two further benefits. 
The first benefit is that the property of two string diagrams being equal if and only if they ``look the same'' can be spelled out as graph isomorphism, eliminating any possible ambiguity of interpretation. 
The second benefit is that equational reasoning can also be formalised as (categorical) graph rewriting, for instance double-pushout rewriting (DPO).

Finally, to represent diagrammatically two concepts essential to both programming languages (binding) and to e-graphs (equivalence classes) we rely on the concept of \emph{functorial boxes} as introduced by Melli\`es~\cite{https://link.springer.com/chapter/10.1007/11874683_1}. 
The diagrammatic boxes require the introduction of further structure to (hyper)graphs in the form of a hierarchical layering of the graphs, as studied in \emph{loc. cit.}.

Revisiting the simple earlier example $\lambda f . f ((\lambda x . f x) 2)$, note its string diagrammatic representation below, in which the two occurrences of $f$ are now shared.
$\lambda$-abstraction is represented by a rounded box with the bound variable a dangling wire attached to the frame,
the application node a half-circle
\adjustbox{scale=0.15}{\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=none] (0) at (1, 6.75) {};
		\node [style=none] (1) at (1.25, 7.75) {};
		\node [style=none] (2) at (2, 8) {};
		\node [style=none] (3) at (2.75, 7.75) {};
		\node [style=none] (4) at (3, 6.75) {};
		\node [style=none] (5) at (1.5, 6.75) {};
		\node [style=none] (6) at (2.5, 6.75) {};
		\node [style=none] (7) at (2, 9) {};
		\node [style=none] (8) at (1.5, 5.75) {};
		\node [style=none] (9) at (2.5, 5.75) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=lambda_unit_string] (3.center)
			 to [in=60, out=-30, looseness=0.75] (4.center)
			 to (0.center)
			 to [in=-150, out=120, looseness=0.75] (1.center)
			 to [in=-180, out=45, looseness=0.75] (2.center)
			 to [in=135, out=0, looseness=0.75] cycle;
		\draw (2.center) to (7.center);
		\draw (5.center) to (8.center);
		\draw (9.center) to (6.center);
	\end{pgfonlayer}
\end{tikzpicture}
}, 
and sharing a \emph{contraction node} DRG: IT'S UPSIDE DOWN?
\scalebox{0.3}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

DRG USE WRAP-BOX. 
\[
\adjustbox{scale=0.6}{
    \tikzfig{figures/de-brujin-string}
}
\]

@@

Nodes of an e-graph are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs rooted at these boxes.
A term graph is embedded as an e-graph by making each equivalence class have a single node.
The first rewrite rule, replacing multiplication by \emph{shift-left} creates the first non-unitary equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) nodes,  now sharing the node `$a$'.
This second e-graph illustrates the other representational difference, besides sharing, between conventional e-graphs and e-hypergraphs.
In the former, edges connect directly to nodes inside the equivalence class, whereas in the latter edges connect to the equivalence class itself.
Explicit discarding (weakening) nodes, depicted with black dots with a single wire inside the dashed boxes 
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}, indicate which of the class-level edges are connected to which nodes inside the class.

>>>>

% for quotienting
\usepackage{xfrac} 
\usepackage{wrapfig}

@@

Another distinct challenge is presented by $\beta$-reduction, which is handled using \emph{explicit substitutions}. 
However, calculi of explicit substitutions have many technical subtleties which also directly impact how equivalence is defined, for instance \emph{graph equivalence} instituted between terms which have non-interfering explicit substitutions applied in different orders~\cite{accattoli2014nonstandard}. 
These difficulties are not handled in \emph{loc. cit.}.

@@

The general background literature for string diagrams is summarised in~\cite{piedeleu2023introductionstringdiagramscomputer}.
The particular context of string diagrams with bindings is given in~\cite{ghica2024stringdiagramslambdacalculifunctional}, while string diagrams for e-graphs without bindings are defined in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.

String diagrams provide a geometric or topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out. 
Key structures in monoidal categories---such as the monoidal product, tensor unit, braiding, and dualities---are represented through operations like placing, weaving, or bending strings. 
This visual approach makes complex operations such as the trace intuitive and reveals insights sometimes obscured by symbolic notation. 
Originating from notations by Hotz~\cite{hotzsd} (for automata) and Penrose~\cite{penrose1984spinors} (for tensor networks), string diagrams generalise to any monoidal category and even bicategories, offering a dual perspective to commutative diagrams through Poincaré duality.

As described above, string diagrams are commonly an informal yet rigorous two-dimensional syntax for the internal language of a category. 
They can be formalised by representing them as graphs or hypergraphs~\cite{1011453502719}, which has two further benefits. 
The first benefit is that the property of two string diagrams being equal if and only if they ``look the same'' can be spelled out as graph isomorphism, eliminating any possible ambiguity of interpretation. 
The second benefit is that equational reasoning can also be formalised as (categorical) graph rewriting, for instance double-pushout rewriting (DPO).

Finally, to represent diagrammatically two concepts essential to both programming languages (binding) and to e-graphs (equivalence classes) we rely on the concept of \emph{functorial boxes} as introduced by Melli\`es~\cite{10.1007/11874683_1}. 
The diagrammatic boxes require the introduction of further structure to (hyper)graphs in the form of a hierarchical layering of the graphs, as studied in \emph{loc. cit.}.

\begin{wrapfigure}{L}{0.2\textwidth}
\[
\adjustbox{scale=0.6}{
    \tikzfig{figures/de-brujin-string}
}
\]
\end{wrapfigure}

Revisiting the simple earlier example $\lambda f . f ((\lambda x . f x) 2)$, note its string diagrammatic representation below, in which the two occurrences of $f$ are now shared.
$\lambda$-abstraction is represented by a rounded box with the bound variable a dangling wire attached to the frame,
the application node a half-circle
\adjustbox{scale=0.15}{\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=none] (0) at (1, 6.75) {};
		\node [style=none] (1) at (1.25, 7.75) {};
		\node [style=none] (2) at (2, 8) {};
		\node [style=none] (3) at (2.75, 7.75) {};
		\node [style=none] (4) at (3, 6.75) {};
		\node [style=none] (5) at (1.5, 6.75) {};
		\node [style=none] (6) at (2.5, 6.75) {};
		\node [style=none] (7) at (2, 9) {};
		\node [style=none] (8) at (1.5, 5.75) {};
		\node [style=none] (9) at (2.5, 5.75) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=lambda_unit_string] (3.center)
			 to [in=60, out=-30, looseness=0.75] (4.center)
			 to (0.center)
			 to [in=-150, out=120, looseness=0.75] (1.center)
			 to [in=-180, out=45, looseness=0.75] (2.center)
			 to [in=135, out=0, looseness=0.75] cycle;
		\draw (2.center) to (7.center);
		\draw (5.center) to (8.center);
		\draw (9.center) to (6.center);
	\end{pgfonlayer}
\end{tikzpicture}
}, 
and sharing a \emph{contraction node} 
\scalebox{0.3}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

@@

Nodes of an e-graph are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs rooted at these boxes.
A term graph is embedded as an e-graph by making each equivalence class have a single node.
The first rewrite rule, replacing multiplication by \emph{shift-left} creates the first non-unitary equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) nodes,  now sharing the node `$a$'.
This second e-graph illustrates the other representational difference, besides sharing, between conventional e-graphs and e-hypergraphs.
In the former, edges connect directly to nodes inside the equivalence class, whereas in the latter edges connect to the equivalence class itself.
Explicit discarding (weakening) nodes, depicted with black dots with a single wire inside the dashed boxes

\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}, indicate which of the class-level edges are connected to which nodes inside the class.
